Thesis: draft

### 2021-05-07 draft.tex
latest update

Another important tool in the study of opinion dynamics/formations in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms, molecules,... to macroscopic observed behavior and has a wide field of applications inside and outside the world of physics. The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics or sociophysics. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display stunning regularities, transitions from disorder to order, the emergence of consensus, universality, etc \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the found regularities at large scale as collective effects of the interactions among these individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one big difficulty compared to the description of gases and other physical phenomena. In regular physics, the microscopic constituents are `simple' atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics/behavior of individuals would be well know, then the interactions between them would still be complicated and it would be impossible to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, it may seem that extracting useful and qualitative information out of social models is a nearly impossible and almost hopeless task \cite{Castellano2009}. The use of statistical physics may, however, also offer a way out. In most situations the qualitative (and sometimes even quantitative) properties of large scale phenomena do not depend on the microscopic details of the process \cite{Castellano2009}. Instead, the relevant aspects for the global behavior are often higher level features such as symmetries, conservation laws, dimensionality, etc \cite{Castellano2009}. This introduces a concept of universality. With this universality in mind, one can proceed to build models of human/social behavior where only the simplest and most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with real/empirical data to investigate whether or not the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}. This interplay between theory, model, experiment and real-life data is not only an important factor in social physics, but also in regular physical research and science in general.

@@

The bounded confidence models, discussed in Section \ref{bounded}, can be easily extended to included some notion of stubbornness. Guillaume Deffuant proposed such an extension where stubbornness is correlated to extremism \cite{Deffuant2006}. The model he proposed goes as follows. Remember that in the bounded confidence models, the agents have an opinion $x_i$ (which is a real number between $\interval{-1}{1}$) and a threshold $\epsilon_i$, which determines the range of opinions that can influence agent $i$ (agents with opinions between $\interval{x_i - \epsilon_i}{x_i + \epsilon_i}$ are capable of influencing agent $i$). Guillaume Deffuant then introduced a fraction of extremists, $f_e$, which can have either the extreme opinion $-1$ or the extreme opinion $1$ \cite{Deffuant2006}. These extremists are given a threshold equal to $\epsilon_e$ \cite{Deffuant2006}. The other agents are considered to be moderate and can have opinions between $-1$ and $1$. The moderates are given a threshold $\epsilon$, with $\epsilon > \epsilon_e$ \cite{Deffuant2006}. This last condition implicitly denotes the resistance of the extremes towards other opinions in contrast to the more flexible nature of the moderates, which is based on the idea that people with extreme opinions will be more convinced of that opinion and will be less likely to change it.\\
\newline
Since this thesis is not concerned with bounded confidence models, we will not discuss the results of this model here. Interested readers can find them in reference \cite{Deffuant2006}. \\
\newline
Another example of introducing some kind of persistence towards the own opinion of the nodes is the ``hardening of positions'' that was briefly discussed at the end of Section \ref{classMod}.

@@

This model is a simple one, since it only allows for two different opinions in the system. The designed filtering algorithms are simple and intuitive as well. Some randomness is introduced by using the probabilistic majority model instead of the classic majority model. This might make the model more realistic. This simpleness combined with the touch of randomness to give the model a more realistic vibe are definitely some strong points of using this model compared to other, more complex models. A simple model is more tractable and allows for a better understanding of what is going on in the dynamics. However, this might be a downside as well. Real-world opinion dynamics are, probably, far from simple. People do not update their opinion according to a simple probabilistic opinion dynamics model. Beside that, the cases where people have to choose between only two opinions are rather limited. Filtering algorithms, used by real social media companies, are unfortunately not publicly shared. We can however assume that they are not as simple as the ones we used in this thesis. Trying to model these far more complex systems by our very simple model, might lead to the fact that a lot of the dynamics might be missed or not well captured by our model. This downside of trying to model real-world social behavior with simple, tractable models is really a call for the urgent need for more real-world data to which the results obtained from simple models can be compared and which can help make the models capable of generating more realistic results. \\

@@

This does not mean that simple models should be thrown away. They might still be capable of capturing a lot of the dynamics that are seen in the real world. Not because a system appears complex, that it cannot be modeled by something far more simple. This has been proven by physics research over and over again. Beside that, if you do not know the dynamics/underlying processes of complex phenomena yet, the best thing you can do is try to model it with the most simple model that comes to mind. There is no need to make things too complex if simple works too. There is however a need to compare data from (computational) models to experimental, real-world data. A need for critical thinking and careful analyzing of results can also not be denied.

@@

The formation of echo chambers versus the fraction of completely stubborn actors, fracRes, in the Last.fm on-line social network is studied. The results are compared to corresponding results from network models, all with $N = 8003$: (i) the Watts-Strogatz model with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) the stochastic block model with 53 communities of size 151, $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$; (iii) the stochastic Watts-Strogatz block model with 53 communities of size 151, each community has the following Watts-Strogatz parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$. The results for the PR and REC filtering algorithm are shown in Fig. \ref{echo_vs_fracRes_real}.\\
\newline
The PR filtering algorithm shows a decrease in the formation of echo chambers with an increasing fraction of stubborn actors. This is valid for both the real-world network and the network models. The stochastic Watts-Strogatz block model has a similar clustering and community structure as the Last.fm network. The Last.fm network has, however, a smaller formation of echo chambers than the SBM-WS. This suggests that it is not only the clustering and/or the community structure that drives the formation of echo chambers. Other topologies must play a role as well. The biggest difference between the real-world Last.fm network and the stochastic Watts-Strogatz block model lies in the degree distribution: the Last.fm network has a heterogeneous degree distribution instead of a homogeneous. It is probably this heterogeneous contact pattern that hampers the formation of echo chambers in the real-world network compared to a network model with similar clustering and modularity, but with a homogeneous degree distribution.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{REC}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_PR_real_network_8x7.png}
    \label{pr_real}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{PR}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_REC_real_network_8x7.png}
    \label{rec_real}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of completely stubborn actors for the PR and REC filtering algorithms and an initial $50/50$ opinion distribution. The Last.fm on-line social network is compared to corresponding network models.]{Formation of echo chambers versus the fraction of completely stubborn actors, fracRes, for an initial $50/50$ opinion distribution. The Last.fm on-line social network is compared to corresponding network models: (i) WS with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) SBM-WS with 53 communities of size 151, each community has the following Watts-Strogatz parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$; (iii) SBM with 53 communities of size 151 with $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. The $y$-axis shows the average of the normalized fraction of nodes with all neighbors having opinion A and the normalized fraction of nodes with all neighbors having opinion B. This value represents the formation of echo chambers in the system (denoted with echo chamber size in the figure). Each data point is the average of $50$ independent simulations. Bars represent the standard deviation $\sigma$.}
\label{echo_vs_fracRes_real}
\end{figure}

@@

We investigated the effect of the network structure on the dynamics of the opinions in the system. The main focus was put on the difference between networks with a high clustering coefficient and no community structure (WS) compared to networks with a clear community structure and a low clustering coefficient (SBM). It was found that a clear community structure does enhance the formation of echo chambers (mainly when using the PR filtering algorithm) compared to a random network. This enhancement is however not as strong as for networks with a high clustering coefficient. Networks that have both a high clustering coefficient and a high modularity (clear community structure) (SBM-WS) behave similar as the high clustering networks (WS). This indicates that it is mainly the high clustering that drives the formation of echo chambers. There is a correlation between modularity and formation of echo chambers: a higher modularity results in a larger formation of echo chambers (in case of the PR filtering algorithm). The enhancement is just not as strong as the enhancement in the formation of echo chambers introduced by a high clustering. \\

@@

When introducing a notion of stubbornness (with the help of a stubbornness parameter $r$), we found that the interplay between the PR filtering algorithm and this individual resistance resulted in a decrease in the formation of echo chambers for the both the high clustering (WS, SBM-WS) and the high modularity networks (SBM, SBM-WS). The REF and REC filtering algorithms, on the other hand, showed an increase in the formation of echo chambers with increasing stubbornness, followed by a decrease if the stubbornness was further increased. This is mainly seen in the high clustering networks, however a small bump can also be observed in the high modularity network. This interplay between the PR filtering algorithm and individual resistance versus the interplay between the REF/REC filtering algorithms and stubbornness is seen for both balanced starting conditions and unbalanced starting conditions.\\

@@

It was found that the way the opinions are initially distributed over the communities in a network with a community structure (SBM) does make a difference. We investigated a scenario in which a fraction of the communities was given a single opinion, opinion A, while the remaining communities were given an initial $50/50$ opinion distribution. This scenario was then compared to the case in which the total opinion distribution was randomly distributed over the system. This was done for both a high modularity and a low modularity network and for both the REC and PR filtering algorithms. \\

@@

The formation of echo chambers, and the enhancement of them when using algorithmic personalization such as semantic filtering, agrees with results found in literature \cite{Cinelli2020}\cite{Nikolov2015}. The enhancement of the formation of echo chambers in clustered communities is also found in literature on complex contagious processes \cite{Becker2017}.\\

@@

The hampering in the increase of the majority opinion in case of unbalanced starting conditions combined with the introduction of stubborn actors for the PR filtering algorithm agrees partly with related research on continuous opinion dynamics models combined with stubborn actors \cite{Como2013}. In this research, Acemo\u{g}lu et al. showed that introducing stubborn actors (with different opinions) in an inhomogeneous stochastic gossip process of continuous opinion dynamics leads to a situation in which consensus can no longer be reached \cite{Como2013}. This result is also found in the majority opinion dynamics model and in the voter model combined with stubborn actors (equally distributed over the two opinions) \cite{Galam2007}\cite{Mobilia2007}\cite{Yildiz2013}. In our case, the possible consensus around the majority opinion, that could have been reached when combining the PR filtering algorithm with unbalanced starting conditions, can no longer be obtained when introducing stubborn actors.

>>>>

Another important tool in the study of opinion dynamics/formations in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms, molecules,... to macroscopic observed behavior and has a wide field of applications inside and outside the world of physics. The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics or sociophysics. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display stunning regularities, transitions from disorder to order, the emergence of consensus, universality, etc \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the found regularities at large scale as collective effects of the interactions among these individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one big difficulty compared to the description of gases and other physical phenomena. In regular physics, the microscopic constituents are `simple' atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics/behavior of individuals would be well know, then the interactions between them would still be complicated and it would be impossible to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, it may seem that extracting useful and qualitative information out of social models is a nearly impossible and almost hopeless task \cite{Castellano2009}. The use of statistical physics may, however, also offer a way out. In most situations the qualitative (and sometimes even quantitative) properties of large scale phenomena do not depend on the microscopic details of the process \cite{Castellano2009}. Instead, the relevant aspects for the global behavior are often higher level features such as symmetries, conservation laws, dimensionality, etc \cite{Castellano2009}. This introduces a concept of universality. With this universality in mind, one can proceed to build models of human/social behavior where only the simplest and most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with real/empirical data to investigate whether or not the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}\cite{Pawel}. This interplay between theory, model, experiment and real-life data is not only an important factor in social physics, but also in regular physical research and science in general.

@@

The bounded confidence models, discussed in Section \ref{bounded}, can be easily extended to included some notion of stubbornness. Guillaume Deffuant proposed such an extension where stubbornness is correlated to extremism \cite{Deffuant2006}. The model he proposed goes as follows. Remember that in the bounded confidence models, the agents have an opinion $x_i$ (which is a real number between $\interval{-1}{1}$) and a threshold $\epsilon_i$, which determines the range of opinions that can influence agent $i$ (agents with opinions between $\interval{x_i - \epsilon_i}{x_i + \epsilon_i}$ are capable of influencing agent $i$). Guillaume Deffuant then introduced a fraction of extremists, $f_e$, which can have either the extreme opinion $-1$ or the extreme opinion $1$ \cite{Deffuant2006}. These extremists are given a threshold equal to $\epsilon_e$ \cite{Deffuant2006}. The other agents are considered to be moderate and can have opinions between $-1$ and $1$. The moderates are given a threshold $\epsilon$, with $\epsilon > \epsilon_e$ \cite{Deffuant2006}. This last condition implicitly denotes the resistance of the extremes towards other opinions in contrast to the more flexible nature of the moderates, which is based on the idea that people with extreme opinions will be more convinced of that opinion and will be less likely to change it.\\
\newline
Since this thesis is not concerned with bounded confidence models, we will not discuss the results of this model here. Interested readers can find them in reference \cite{Deffuant2006}. \\
\newline
Another example of introducing some kind of persistence towards the own opinion of the nodes is the ``hardening of positions'' that was briefly discussed at the end of Section \ref{classMod}. The model introduced by Friedkin and Johnsen in which each agent adheres to its initial opinion to a certain degree, see Section \ref{classMod}, can also be seen as introducing a notion of stubbornness \cite{Noorazar2020}.

@@

This model is a simple one, since it only allows for two different opinions in the system. The designed filtering algorithms are simple and intuitive as well. Some randomness is introduced by using the probabilistic majority model instead of the classic majority model. This might make the model more realistic. This simpleness combined with the touch of randomness to give the model a more realistic vibe are definitely some strong points of using this model compared to other, more complex models. A simple model is more tractable and allows for a better understanding of what is going on in the dynamics. However, this might be a downside as well. Real-world opinion dynamics are, probably, far from simple. People do not update their opinion according to a simple probabilistic opinion dynamics model. Beside that, the cases where people have to choose between only two opinions are rather limited. Filtering algorithms, used by real social media companies, are unfortunately not publicly shared. We can however assume that they are not as simple as the ones we used in this thesis. Trying to model these far more complex systems by our very simple model, might lead to the fact that a lot of the dynamics might be missed or not well captured by our model. This downside of trying to model real-world social behavior with simple, tractable models is really a call for the urgent need for more real-world data to which the results obtained from simple models can be compared and which can help make the models capable of generating more realistic results \cite{Pawel}. \\

@@

This does not mean that simple models should be thrown away. They might still be capable of capturing a lot of the dynamics that are seen in the real world. Not because a system appears complex, that it cannot be modeled by something far more simple. This has been proven by physics research over and over again. Beside that, if you do not know the dynamics/underlying processes of complex phenomena yet, the best thing you can do is try to model it with the most simple model that comes to mind. There is no need to make things too complex if simple works too. There is however a need to compare data from (computational) models to experimental, real-world data. A need for critical thinking and careful analyzing of results can also not be denied \cite{Pawel}.

@@

The formation of echo chambers versus the fraction of completely stubborn actors, fracRes, in the Last.fm on-line social network is studied. The results are compared to corresponding results from network models, all with $N = 8003$: (i) the Watts-Strogatz model with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) the stochastic block model with 53 communities of size 151, $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$; (iii) the stochastic Watts-Strogatz block model with 53 communities of size 151, each community has the following Watts-Strogatz parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$. The results for the PR and REC filtering algorithm are shown in Fig. \ref{echo_vs_fracRes_real}.\\
\newline
The PR filtering algorithm shows a decrease in the formation of echo chambers with an increasing fraction of stubborn actors. This is valid for both the real-world network and the network models. The stochastic Watts-Strogatz block model has a similar clustering and community structure as the Last.fm network. The Last.fm network has, however, a smaller formation of echo chambers than the SBM-WS. This suggests that it is not only clustering and/or community structure that drives the formation of echo chambers. Other topologies must play a role as well. The biggest difference between the real-world Last.fm network and the stochastic Watts-Strogatz block model lies in the degree distribution: the Last.fm network has a heterogeneous degree distribution instead of a homogeneous degree distribution. It is probably this heterogeneous contact pattern that hampers the formation of echo chambers in the real-world network compared to a network model with similar clustering and modularity, but with a homogeneous degree distribution. This agrees with the result that Perra and Rocha \cite{Perra2019} found: a heterogeneous contact pattern hampers the formation of echo chambers. % talk about possible effect of powerlaw community sizes + figure??
The results for the Watts-Strogatz model are similar to those for the stochastic Watts-Strogatz block model. The formation of echo chambers seems to be slightly smaller in the WS than in the SBM-WS network. It may be tempting to attribute this to the lack of community structure in the WS model. It must however be noted that the implemented WS model has a slightly smaller clustering coefficient ($\left<cc\right> = 0.317 \pm 0.004$) than the SBM-WS ($\left<cc\right> = 0.344 \pm 0.002$), which may also explain the somewhat smaller formation of echo chambers in the WS network than in the SBM-WS network. Furthermore, it is interesting to notice that the Last.fm on-line social network coincides with the SBM with a similar modularity. It seems that the hampering of the formation of echo chambers due to a heterogeneous contact pattern is of the same size as the hampering due to a lack of high clustering. % talk about coincidence??

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{REC}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_PR_real_network_8x7.png}
    \label{pr_real}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{PR}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_REC_real_network_8x7.png}
    \label{rec_real}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of completely stubborn actors for the PR and REC filtering algorithms and an initial $50/50$ opinion distribution. The Last.fm on-line social network is compared to corresponding network models.]{Formation of echo chambers versus the fraction of completely stubborn actors, fracRes, for an initial $50/50$ opinion distribution. The Last.fm on-line social network is compared to corresponding network models: (i) WS with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) SBM-WS with 53 communities of size 151, each community has the following Watts-Strogatz parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$; (iii) SBM with 53 communities of size 151 with $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. The $y$-axis shows the average of the normalized fraction of nodes with all neighbors having opinion A and the normalized fraction of nodes with all neighbors having opinion B. This value represents the formation of echo chambers in the system (denoted with echo chamber size in the figure). Each data point is the average of $50$ independent simulations. Bars represent the standard deviation $\sigma$.}
\label{echo_vs_fracRes_real}
\end{figure}
\noindent
The REC filtering algorithm enforces an increase in the formation of echo chambers when a small fraction of stubborn actors is introduced. If this fraction is further increased, the formation of echo chambers starts to decrease again. The increase is the strongest for the high clustering network models (WS, SBM-WS). Even though the Last.fm network has a similar clustering as these high clustering models, the formation of echo chambers and the corresponding increase in that formation are significantly smaller. As for the PR filtering algorithm, this is probably a consequence of the heterogeneous degree pattern of the Last.fm network. % talk about SBM

@@

We investigated the effect of the network structure on the dynamics of the opinions in the system. The main focus was put on the difference between networks with a high clustering coefficient and no community structure (WS) compared to networks with a clear community structure and a low clustering coefficient (SBM). It was found that a clear community structure enhances the formation of echo chambers (mainly when using the PR filtering algorithm) compared to a random network. This enhancement is however not as strong as for networks with a high clustering coefficient. Networks that have both a high clustering coefficient and a high modularity (clear community structure) (SBM-WS) behave similar as the high clustering networks (WS). This indicates that it is mainly the high clustering that drives the formation of echo chambers. There is a correlation between modularity and formation of echo chambers: a higher modularity results in a larger formation of echo chambers (in case of the PR filtering algorithm). The enhancement is just not as strong as the enhancement in the formation of echo chambers introduced by a high clustering. \\

@@

When introducing a notion of stubbornness (with the help of a stubbornness parameter $r$), we found that the interplay between the PR filtering algorithm and this individual resistance resulted in a decrease in the formation of echo chambers for the both the high clustering (WS, SBM-WS) and the high modularity networks (SBM, SBM-WS). The REF and REC filtering algorithms, on the other hand, showed an increase in the formation of echo chambers with increasing stubbornness, followed by a decrease if the stubbornness was further increased. This is mainly seen in the high clustering networks, however a small bump can also be observed in the high modularity network. This interplay between the PR filtering algorithm and individual resistance versus the interplay between the REF/REC filtering algorithms and stubbornness is seen for both balanced starting conditions and unbalanced starting conditions. It was also found that including stubborn actors hampers the increase in the majority opinion (and thus prevents the formation of a final consensus state) in case of unbalanced starting conditions combined with the PR filtering algorithm.\\

@@

Finally, it was found that the initial distribution of the opinions over the communities in a network with community structure (SBM) does make a difference. We investigated a scenario in which a fraction of the communities was given a single opinion, opinion A, while the remaining communities were given an initial $50/50$ opinion distribution. This scenario was then compared to the case in which the total opinion distribution was randomly distributed over the system. This was done for both a high modularity and a low modularity network and for both the REC and PR filtering algorithms. \\

@@

The formation of echo chambers in social networks/social media, and the enhancement of them when using algorithmic personalization such as semantic filtering, agrees with results found in literature \cite{Baumann2019}\cite{Cinelli2020}\cite{Nikolov2015}\cite{Sirbu2018}. The enhancement of the formation of echo chambers in clustered communities is also found in literature on complex contagious processes \cite{Becker2017}. Banisch and Olbrich showed that in a social feedback model polarization is increased in high modularity networks compared to an Erd\H{o}s-R\'{e}nyi network \cite{Banisch2018}. This agrees with our findings that a clear community structure enhances the formation of echo chambers.\\

@@

The hampering in the increase of the majority opinion in case of unbalanced starting conditions combined with the introduction of stubborn actors for the PR filtering algorithm agrees partly with related research on continuous opinion dynamics models combined with stubborn actors \cite{Como2013}. In this research, Acemo\u{g}lu et al. showed that introducing stubborn actors (with different opinions) in an inhomogeneous stochastic gossip process of continuous opinion dynamics leads to a situation in which consensus can no longer be reached \cite{Como2013}. This result is also found in the majority opinion dynamics model and in the voter model combined with stubborn actors (equally distributed over the two opinions) \cite{Galam2007}\cite{Mobilia2007}\cite{Yildiz2013}. In our case, the possible consensus around the majority opinion, that could have been reached when combining the PR filtering algorithm with unbalanced starting conditions, can no longer be obtained when introducing stubborn actors.\\
Wang et al. \cite{Wang2020} proposed an agent based continuous opinion dynamics model in which they investigated the formation of echo chambers. They found that stubbornness of individuals protects them from the
influence of centralized sources of influence; ``A purely stubborn or conformist
population may both be less susceptible to the emergence of
two dominant and opposing echo chambers'' (Wang et al., p.10, \cite{Wang2020}). This agrees with our result that the PR filtering algorithm, combined with stubborn actors, hinders the formation of echo chambers.
